The interactions of rational, pragmatic agents\\ lead to efficient language structure and use

### 2019-01-31 paper/paper.tex
updates Thurs morning.

Zipf articulated what is now considered a \emph{functionalist} approach
to language science -- analyzing language structure and use in terms of
efficiency. Such an approach might reframe our opening question as
follows: how does having property \textit{x} make using language
\(\ell\) more or less useful for communication? Among others, this
efficiency-based framing has produced a rich set of theoretical and
empirical targets exploring semantic typology (Regier, Kemp, \& Kay,
2015), properties such as ambiguity (Piantadosi, Tily, \& Gibson, 2011)
and compositionality (Kirby, Griffiths, \& Smith, 2014), and the
efficient use of reduction and redundancy in production (Genzel \&
Charniak, 2002; Levy \& Jaeger, 2007). These projects typically employ
information-theoretic principles to characterize efficiency in natural
or simulated language corpora. We adopt a different approach in this
work, deriving a measure of efficiency from first principles of
\textit{efficient language use} and then showing how this can lead to
communicatively efficient systems.\par

Functionalist theories commonly frame language efficiency in terms of a
fundamental effort-assymetry underlying everyday communication: what is
``hard'' for a speaker is likely different than what is ``hard'' for a
listener. Zipf gave the following intution for how this idea -- purely
from the standpoint of speaker effort, an optimal language
\(\ell_{speaker}^*\) would tend toward a vocabulary of a single,
low-cost word. Given such a language, the full set of potential meanings
would be conveyed using only that word, i.e. \(\ell_{speaker}^*\) would
be fully ambiguous and all possible meanings would need to be
disambiguated by a listener. From the standpoint of listener effort, an
optimal language \(\ell_{listener}^*\) would map all possible meanings
to distinct words, removing a listener's need to disambiguate. In this
example, speaker effort is related to \emph{production cost} and
listener effort to \emph{understanding or disambiguation cost}. Clearly,
natural languages fall between the two extremes of \(\ell_{speaker}^*\)
and \(\ell_{listener}^*\). Zipf proposed that the particular
lexicon-level properties he observed were a result of these competing
forces -- the pressure to jointly minimize speaker and listener
effort.\par

@@

In the current work we provide a case-study for the framing presented
above. We choose a property of languages that could, in principle, vary
freely, but shows strong regularities across languages. The explanadum
is why this regularity holds. We examine ambiguity as our property,
extending ideas by Piantadosi et al. (2011). We define a novel measure
of efficiency that depends on the interactional behavior of speaker and
listener agents. We adopt the reference game as our primary unit of
interaction and model language users using the Rational Speech Act (RSA)
framework -- a computional model of language use, which is nicely
attested by experimental data on interaction. Using these basic
ingredients, we show that the property of interest (ambiguity) is
prevalent in languages that optimize our measure of efficiency. Further,
we show how ambiguity is \textit{used} efficiently during local,
in-the-moment interactions. The contributions of this work are twofold
-- we derive a novel objective for measuring linguistic efficiency and
also show how the reference game framework, in combination with formal
models of communication, can be used to connect ideas about system-level
efficiencies to in-the-moment language use.\par

@@

Zipf's example of optimal speaker- and listener-languages took the form
of a reference game. We adopt that formulation here, assuming these
communication games as our basic unit of analysis. In this framework,
speakers and listeners are aware of a set of objects \(M\)
(\emph{meanings}) and are knowledgeable about the set of possible
signals \(U\) (\emph{utterances}) that can be used to refer to a given
meaning (see Figure 1). Utterances may have different relative costs,
operationalized via a prior over utterances \(P(U)\). Similarly,
meanings differ in the relative degree to which they need to be talked
about, operationalized as a prior over meanings
\(P(M)\)\footnote{The prior over meanings are analogous to the \textit{need probabilities} assumed in previous work (Regier, Kemp \& Kay (2015).}.
We consider a set of contexts \(c\in C\) with an associated prior
\(P(C)\). Each context \(c_i\) describes a different distribution over
meanings e.g. \(p(M|C=c_i) \neq p(M|C=c_j)\). Finally, we consider a set
of communicative events \(e \in E\) where \(<u, m, c> = e\) is an
utterance, meaning and context triple.\par

@@

The Rational Speech Act framework (RSA) is computational-level theory of
pragmatic language use, which has been well attested in experimental
work examining both speaker and listener communication behavior (Frank
\& Goodman, 2012; Goodman \& Frank, 2016). RSA is a formalization of
essential Gricean pragmatic principles -- agents reason about one
another and their shared context (Grice, 1975). We adopt RSA as our
representational framework to model Gricean (rational and pragmatic)
speaker-listeners in the reference game
setting\footnote{See SI section 1 for a more detailed description of basic RSA speaker-listener defintions and to Goodman \& Frank [-@GoodmanFrank2016a] for an overview of its applications.}.\par

@@

From an information-theoretic perspective this objective is intuitive:
The Cross-Entropy (CE) gives us a measure of dissimilarity between two
distributions -- the average number of bits required to communicate
under one distribution, given that the ``true'' distribution differs. In
our case, we have an expectation over this term -- the expected
difference between the distributions assumed by the speaker
\(P_{speaker}\) and listener \(P_{listener}\) given a set of contexts
\(C\). In other words, an ``efficient'' language \(\ell\) minimizes the
distance between what speakers and listeners think.

@@

The task of understanding language is marked by a frequent need to
handle various forms of ambiguity: lexical, syntactic, among
others(Wasow, Perfors, \& Beaver, 2005). The ubiquity of this property,
however, has been argued to provide evidence that langauges are not been
optimized for
communication\footnote{Chomsky's (2002) claim that "... language is not well designed, because you have such properties as ambiguity" is analogous to the Zipfian optimal Listener language $\ell_{listener}^*$.}
(Chomsky, 2002).

@@

We re-iterate their argument here. Say we have two objects (\(m_1\) and
\(m_2\)), two utterances (\(u_1\) and \(u_2\)), which differ in cost,
and two languages (\(\ell_1\) and \(\ell_2\)) which describe different
utterance-meaning mappings. In language \(\ell_1\), the low-cost \(u_1\)
can be used to refer to both \(m_1\) and \(m_2\)
(\([[u_1]]_{\ell_1} = \{m_1, m_2 \}\)), but the high-cost \(u_2\) cannot
be used at all (\([[u_2]]_{\ell_1} = \emptyset\)). By contrast, in
language \(\ell_2\), \(u_1\) can only refer to \(m_1\) and \(u_2\) can
only refer to \(m_2\) (\([[u_1]]_{\ell_2} = \{m_1\}\) and
\([[u_2]]_{\ell_2} = \{m_2 \}\)). While it is cheaper for a speaker to
use \(\ell_1\) (because speaking is always low-cost), it is more
difficult for a listener (because \(u_1\) is ambiguous). Crucially,
however, if context is disambiguating then the speaker can use \(u_1\)
to refer to either \(m_1\) or \(m_2\) and \(\ell_1\) should be preferred
to \(\ell_2\).\par

In the following experiments we explore two aspects of Piantadosi's
claim. First, we examine the efficient language \emph{structure} aspect
of their claim, exploring when the optimal linguistic system \(\ell^*\)
is most likely to contain ambiguous lexical items. In the second
experiment, we explore an efficient language \emph{use} aspect of the
claim -- at what point in a conversation is it useful for a speaker to
use ambiguous lexical material?\par

@@

We adopt the hypothesis presented by Piantadosi, formalizing their
argument in our current framework -- showing that ambiguity is an
efficient property under our CE objective in the reference game setting.
We proceed by generating languages with different amounts of contextual
support (varying the size of \(|C|\)). We search the space of languages,
examining whether ones which minimize our objective contain ambiguity.
If context leads to more efficient commnication, then optimal languages
under our objective should be more likley to be ambiguous as the amount
of context increases.\par

@@

We conduct \(N=2000\) simulations. For each simulation we enumerate the
set of \emph{valid} languages in which \(|U|=|M|=4\) (recall that \(U\)
is our set of utterances and \(M\) our set of meanings). Note that a
language \(\ell \in L\) is ``valid'' so long as each possible meaning in
\(m \in M\) can be referred to by at least one form \(u \in U\) (every
column of \(\ell\) has some non-zero assignment) and each form maps to
at least one meaning (every column has some non-zero assignment). For a
given simulation the goal is to find the language \(\ell^*\) which
minimizes our CE objective and then check to see if that language
contains ambiguity.\par

@@

Following the argument given by Piantadosi et al. (2011), we want to
assess the impact of \emph{context} on the presence of ambiguity in
optimal languages. To do so we consider four conditions with \(n=500\)
simulations each (that is, 500 unique sets of
\(\{P(U), P(M|C), P(C)\}\). Our first is a \emph{one-context} condition
(\(|C|=1\)) -- only a single distribution over meanings \(P(M)\). In our
two-context condition (\(|C| = 2\)), we consider efficiency under both
\(P(M|C=c_1)\) as well as \(P(M|C=c_2)\). Three- and four-context
conditions corresponding accordingly.\par

@@

Figure 1. plots the proportion of optimal languages under each objective
as a function of context size. The red line shows that as the number of
contexts increases, so does the probability that an optimal language
\(\ell^*_{cross}\) contains ambiguity under our CE objective. For
comparison we also plot the proportion of speaker-optimal
\(\ell^*_{speaker}\) (blue line) and listener-optimal
\(\ell^*_{listener}\) (green line) languages that contain ambiguity. In
line with Zipf's predictions, if languages are designed only to minimize
speaker effort then optimal languages always contain ambiguity. If
languages are designed to minimize listener effort then ambiguity is
always avoided.\par

In simulation 1 we explored the degree to which ambiguity was an
efficient property when communication was contextualized. We showed that
as contextual information increased, the optimal languages under our CE
objective were more likely to have ambiguous lexical items. Importantly,
we did not see this effect under alternative objectives which only
emphasized speaker- or listener effort alone.\par

While our results indicate that ambiguity is an efficient property of
contextualized language, these simulations assumed that agents could
always disambiguate items in context. That is, in our four conditions
both agents had perfect knowledge of the relevant conditional
distributions (\(P(M|C)\)). This assumption may be too strong for
describing much of day-to-day communication -- we seldom interact with
others with perfect knowledge of the current context (or topic) of
conversation. To explore how ambiguity may be \textit{used} efficiencly,
we next explore a case in which the listener has imperfect knowledge of
context at the start of the conversation, but may infer it from a
discourse history in Simulation 2.\par

@@

We shift focus in Simulation 2 -- instead of examining the landscape of
a space of languages under our CE objective, we consider a single
language containing ambiguity, examining how that property is used over
a discourse. If context is not informative with respect to meaning early
in a discourse, but is at later positions, then an efficient speaking
strategy would avoid ambiguous material early, saving it for when
contextual information is disambiguating. In this toy example we can
consider ``context'' as analogous to a ``topic'' of conversation. As an
example, imagine a scenario in which a reader is beginning a magazine
article. While they may have some knowledge about the article's topic
(perhaps from the title), they may not have complete knowledge of its
contents, including the persons or events involved. In this setting,
using an potential low-cost, but ambiguous pronoun early may lead to
misunderstanding if context is sufficiently informative. But, if by a
later position enough contextual information has accumulated, it may be
efficient for the writer to use the ambiguous material.\par

@@

Over the course of a discourse \(D\), the listener tries to infer both
the current topic, \(c_{current}\), as well as the particular meaning
\(m\) of a given utterance \(u\). That is, we consider updated listener
and speakers (\(L(m, c|u,D;\ell)\), \(S(u|m,c,D;\ell)\)) in which both
agents can track the history of previous utterances \(D\). Importantly,
an agent can attempt to infer the current topic of conversation
\(c_{current}\) using the discourse history \(D\) (see SI section 3 for
more detail).\par

@@

We consider three types of speaker models. Our \textit{Full pragmatics}
agent, models a speaker who reasons about her listener and also has
complete recall of the set of utterances in the discourse \(D\). This
speaker believes that the listener may not know the current topic
\(c_{current}\) at the start of the discrouse, but should infer it over
the discourse. We compare two baselines models. The first, a
\textit{Partial pragmatics} baseline describes a speaker who reasons
about a listener, but assumes they have no access to discourse history.
The second, a \textit{No pragmatics} speaker who does not consider a
listener at all, but produces utterances according to the underlying
langauge semantics (\(\ell\)) and topic probabilities
(\(p(M|C=c_{current}\)) (see SI section 3 for details).\par

@@

Figure 2, (A) shows the empirical probability that a speaker uses an
ambiguous utterance as a function of discourse position. The
\textit{No pragmatic} baseline uses ambiguous utterances frequently and
at a constant rate over the discourse and the \textit{Partial pragmatic}
baseline avoids ambiguous utterances entirely. But, the
\textit{Full pragmatic} model avoids ambiguous material only at the
start of the discourse, employing it increasingly as the discourse
proceeds. (C) tracks our CE objective for each model over the discourse.
Note that the objective decreases for all three models, primarily driven
by the listener updating his belief about the actual topic
(\(P(C=c_{current}|D)\)). However, the objective declines more quickly
under the \textit{Full} and \textit{Partial} pragmatic speakers as
listener agents are better able to infer the correct context.
Additionally, the difference in CE between the \textit{Full}- and
\textit{Partial pragmatic} models at the end of the discourse is driven
by the reduction in speaker costs described in (B). While speaker effort
remains constant in both \textit{No pragmatic} and
\textit{Partial pragmatic} baselines, effort declines in the
\text{Full Pragmatic} model as she increasingly relies on ambiguous
material later in the discourse.\par

>>>>

Zipf articulated what is now considered a \emph{functionalist} approach
to language science -- analyzing language structure and use in terms of
efficiency. Such an approach might reframe our opening question as
follows: how does having property \textit{x} make using language
\(\ell\) more or less useful for communication? Among others, this
efficiency-based framing has produced a rich set of theoretical and
empirical targets exploring semantic typology (Regier, Kemp, \& Kay,
2015), properties such as ambiguity (Piantadosi, Tily, \& Gibson, 2011)
and compositionality (Kirby, Griffiths, \& Smith, 2014), and the
efficient use of reduction and redundancy in production (Genzel \&
Charniak, 2002; Levy \& Jaeger, 2007). These projects typically employ
information-theoretic principles to characterize efficiency in natural
or simulated language corpora. We adopt a different approach in this
work, deriving a measure of efficiency from first principles of
\textit{efficient language use} and show how this can lead to
communicatively efficient systems.\par

Functionalist theories commonly frame language efficiency in terms of a
fundamental effort-assymetry underlying everyday communication: what is
``hard'' for a speaker is likely different than what is ``hard'' for a
listener. Zipf described this as follows -- purely from the standpoint
of speaker effort, an optimal language \(\ell_{speaker}^*\) would tend
toward a vocabulary of a single, low-cost word. Given such a language,
the full set of potential meanings would be conveyed using only that
word, i.e. \(\ell_{speaker}^*\) would be fully ambiguous and all
possible meanings would need to be disambiguated by a listener. From the
standpoint of listener effort, an optimal language \(\ell_{listener}^*\)
would map all possible meanings to distinct words, removing a listener's
need to disambiguate. In this example, speaker effort is related to
\emph{production cost} and listener effort to \emph{understanding or
disambiguation cost}. Clearly, natural languages fall between the two
extremes of \(\ell_{speaker}^*\) and \(\ell_{listener}^*\). Zipf
proposed that the particular lexicon-level properties he observed were a
result of these competing forces -- the pressure to jointly minimize
speaker and listener effort.\par

@@

In the current work we provide a case-study for the framing presented
above. We choose a property of languages that could, in principle, vary
freely, but shows strong regularities across languages. The explanadum
is why this regularity holds. We examine ambiguity as our property,
extending ideas by Piantadosi et al. (2011). We define a novel measure
of efficiency that depends on the interactional behavior of speaker and
listener agents. We adopt the reference game as our primary unit of
interaction and model language users with the Rational Speech Act (RSA)
framework -- a computional model of language use, which is nicely
attested by experimental data on interaction. Using these basic
ingredients, we show that the property of interest (ambiguity) is
prevalent in languages that optimize our measure of efficiency. Further,
we show how ambiguity is \textit{used} efficiently during local,
in-the-moment interactions. The contributions of this work are twofold
-- we derive a novel objective for measuring linguistic efficiency and
also show how the reference game framework, in combination with formal
models of communication, can be used to connect ideas about system-level
efficiencies to in-the-moment language use.\par

@@

Zipf's example of optimal speaker- and listener-languages took the form
of a reference game. We adopt that formulation here, assuming these
communication games as our basic unit of analysis. In this framework,
speakers and listeners are aware of a set of objects \(M\)
(\emph{meanings}) and are knowledgeable about the set of possible
signals \(U\) (\emph{utterances}) that can be used to refer to a given
meaning (see Figure 1). Utterances may have different relative costs,
operationalized via a prior over utterances \(P(U)\). Similarly,
meanings differ in the relative degree to which they need to be talked
about, operationalized as a prior over meanings
\(P(M)\)\footnote{The prior over meanings are analogous to the \textit{need probabilities} assumed in previous work (Regier, Kemp \& Kay (2015).}.
We consider a set of contexts \(C\) with an associated prior \(P(C)\).
Each context \(c\in C\) describes a different distribution over meanings
e.g. \(p(M|C=c_i) \neq p(M|C=c_j)\). Finally, we consider a set of
communicative events \(e \in E\) where \(<u, m, c> = e\) is an
utterance-meaning-context triple.\par

@@

The Rational Speech Act framework (RSA) is a computational-level theory
of pragmatic language use, which has been well attested in experimental
work examining both speaker and listener communication behavior (Frank
\& Goodman, 2012; Goodman \& Frank, 2016). RSA is a formalization of
essential Gricean pragmatic principles -- agents reason about one
another and their shared context (Grice, 1975). We adopt RSA as our
representational framework to model Gricean (rational and pragmatic)
speaker-listeners in the reference game
setting\footnote{See SI section 1 for a more detailed description of basic RSA speaker-listener defintions and to Goodman \& Frank (2016) for an overview of its applications.}.\par

@@

From an information-theoretic perspective this objective is intuitive:
The Cross-Entropy (CE) gives us a measure of dissimilarity between two
distributions -- the average number of bits required to communicate
under one distribution, given that the ``true'' distribution differs. In
our case, we have an expectation over this term -- the expected
difference between the distributions assumed by the speaker
\(P_{speaker}\) and listener \(P_{listener}\) given a set of contexts
\(C\)\footnote{Note that in the single context case $|C|=1$ this objective is simply the speaker-listener Cross-Entropy over utterance-meaning pairs $H_{cross}[P_{speaker}(u, m; \ell), P_{listener}(u, m; \ell)]$}.
In other words, an ``efficient'' language \(\ell\) minimizes the
distance between what speakers and listeners think.

@@

The task of understanding language is marked by a frequent need to
handle various forms of ambiguity: lexical, syntactic, among others
(Wasow, Perfors, \& Beaver, 2005). The ubiquity of this property,
however, has been argued to provide evidence that languages have not
been optimized for
communication\footnote{Chomsky's (2002) claim that "... language is not well designed, because you have such properties as ambiguity" is analogous to the Zipfian optimal Listener language $\ell_{listener}^*$.}
(Chomsky, 2002).

@@

As an example, say we have two objects (\(m_1\) and \(m_2\)), two
utterances (\(u_1\) and \(u_2\)), which differ in cost, and two
languages (\(\ell_1\) and \(\ell_2\)), which describe different
utterance-meaning mappings. In language \(\ell_1\), the low-cost \(u_1\)
can be used to refer to both \(m_1\) and \(m_2\)
(\([[u_1]]_{\ell_1} = \{m_1, m_2 \}\)), but the high-cost \(u_2\) cannot
be used at all (\([[u_2]]_{\ell_1} = \emptyset\)). By contrast, in
language \(\ell_2\), \(u_1\) can only refer to \(m_1\) and \(u_2\) can
only refer to \(m_2\) (\([[u_1]]_{\ell_2} = \{m_1\}\) and
\([[u_2]]_{\ell_2} = \{m_2 \}\)). While it is cheaper for a speaker to
use \(\ell_1\) (because speaking is always low-cost), it is more
difficult for a listener (because \(u_1\) is ambiguous). Crucially,
however, if context is disambiguating then the speaker can use \(u_1\)
to refer to either \(m_1\) or \(m_2\) and \(\ell_1\) should be preferred
to \(\ell_2\).\par

In the following experiments we explore two aspects of Piantadosi's
claim. In Simulation 1, we examine the efficient language
\emph{structure} aspect of their claim, exploring when the optimal
linguistic system \(\ell^*\) is most likely to contain ambiguous lexical
items. In Simulation 2, we explore an efficient language \emph{use}
aspect of the claim -- at what point in a conversation is it useful for
a speaker to use ambiguous lexical material?\par

@@

We adopt the hypothesis presented by Piantadosi, formalizing their
argument in our current framework -- showing that ambiguity is an
efficient property under our CE objective in the reference game setting.
We proceed by generating languages with different amounts of contextual
support (varying the size of \(|C|\)). We search the space of languages,
examining whether ones which minimize our objective contain ambiguity.
If context leads to more efficient communication, then optimal languages
under our objective should be more likley to be ambiguous as the amount
of context increases.\par

@@

We conduct \(N=2000\) simulations. For each simulation we enumerate the
set of \emph{valid} languages in which \(|U|=|M|=4\) (recall that \(U\)
is our set of utterances and \(M\) our set of meanings). Note that a
language \(\ell \in L\) is ``valid'' so long as each possible meaning in
\(m \in M\) can be referred to by at least one form \(u \in U\) (every
column of \(\ell\) has some non-zero assignment) and each form maps to
at least one meaning (every column has some non-zero assignment). For a
given simulation, the goal is to find the language \(\ell^*\) which
minimizes our objective and then check to see if that language contains
ambiguity.\par

@@

Following the argument given by Piantadosi et al. (2011), we want to
assess the impact of \emph{context} on the presence of ambiguity in
optimal languages. To do so we consider four conditions with \(n=500\)
simulations each (that is, 500 unique sets of
\(\{P(U), P(M|C), P(C)\}\). Our first is a \textit{one-context}
condition (\(|C|=1\)) -- only a single distribution over meanings
\(P(M)\). In our \textit{two-context} condition (\(|C| = 2\)), we
consider efficiency under both \(P(M|C=c_1)\) as well as \(P(M|C=c_2)\).
\textit{Three-} and \textit{four-context} conditions corresponding
accordingly.\par

@@

In Simulation 1 we explored the degree to which ambiguity is an
efficient property of languages when communication is contextualized.
Figure 1. plots the proportion of optimal languages under each objective
as a function of context size. The red line shows that as the number of
contexts increases, so does the probability that an optimal language
\(\ell^*_{cross}\) contains ambiguity under our CE objective. For
comparison we also plot the proportion of speaker-optimal
\(\ell^*_{speaker}\) (blue line) and listener-optimal
\(\ell^*_{listener}\) (green line) languages that contain ambiguity. In
line with Zipf's predictions, if languages are designed only to minimize
speaker effort then optimal languages always contain ambiguity. If
languages are designed to minimize listener effort then ambiguity is
always avoided.\par

While our results indicate that ambiguity is an efficient property of
contextualized language, these simulations assumed that agents could
always disambiguate items in context. That is, in our four conditions
both agents had perfect knowledge of the relevant conditional
distributions (\(P(M|C)\)). This assumption may be too strong for
describing much of day-to-day communication -- we seldom interact with
others with perfect knowledge of the current context (or topic) at the
start of a conversation. To explore how ambiguity may be \textit{used}
efficiently in our framework, we next examine a case in which the
listener has imperfect knowledge of context at the start of the
conversation, but may infer it from the contents of discourse.\par

@@

We shift focus in Simulation 2 -- instead of examining the landscape of
a space of languages under our CE objective, we consider a single
language containing ambiguity, examining how that property is used over
a discourse. If context is not informative with respect to meaning early
in a discourse, but is at later positions, then an efficient speaking
strategy would avoid ambiguous material early, saving it for when
contextual information is disambiguating. We can consider ``context'' as
analogous to a ``topic'' of conversation. As an example, imagine a
scenario in which a reader is beginning a magazine article. While they
may have some knowledge about the article's topic (perhaps from the
title), they may not have complete knowledge of its contents, including
the persons or events involved. In this setting, using a low-cost, but
ambiguous pronoun early may lead to misunderstanding if context is not
informative. But, if by a later position enough contextual information
has accumulated, it may be efficient to use the ambiguous material.\par

@@

Over the course of a discourse \(D\), the listener tries to infer both
the current topic, \(c_{current}\), as well as the particular meaning
\(m\) of a given utterance \(u\). That is, we consider updated listener
and speakers (\(L(m, c|u,D;\ell)\), \(S(u|m,c,D;\ell)\)) in which both
agents can track the history of previous utterances \(D\). Importantly,
an agent can attempt to infer the current topic of conversation
\(c_{current}\) using the discourse history \(D\).\par

@@

We consider three types of speaker models. Our \textit{Full pragmatics}
agent, models a speaker who reasons about her listener and also has
complete recall of the set of utterances in the discourse \(D\). This
speaker believes that the listener may not know the current topic
\(c_{current}\) at the start of the discrouse, but can infer it over the
discourse. We compare two baseline models. The first, a
\textit{Partial pragmatics} baseline describes a speaker who reasons
about a listener, but assumes they have no access to discourse history.
The second, a \textit{No pragmatics} speaker who does not consider a
listener at all, but produces utterances according to the underlying
langauge semantics (\(\ell\)) and topic probabilities
(\(p(M|C=c_{current}\)) (see SI section 3 for details).\par

@@

Figure 2, (A) shows the empirical probability that a speaker uses an
ambiguous utterance as a function of discourse position. The
\textit{No pragmatic} baseline uses ambiguous utterances frequently and
at a constant rate over the discourse and the \textit{Partial pragmatic}
baseline avoids ambiguous utterances entirely. But, the
\textit{Full pragmatic} model avoids ambiguous material only at the
start of the discourse, employing it increasingly as the discourse
proceeds. (C) tracks our CE objective for each model over the discourse.
Note that the objective decreases for all three models, primarily driven
by the listener updating his belief about the actual topic
(\(P(C=c_{current}|D)\)). However, the objective declines more quickly
under the \textit{Full} and \textit{Partial} pragmatic speakers as
listener agents are better able to infer the correct context.
Additionally, the difference in CE between the \textit{Full}- and
\textit{Partial pragmatic} models at the end of the discourse is driven
by the reduction in speaker costs described in (B). While speaker effort
remains constant in both \textit{No pragmatic} and
\textit{Partial pragmatic} baselines, effort declines in the
\textit{Full pragmatic} model as she increasingly relies on ambiguous
material later in the discourse.\par
